---
title: "Aggregate occurrence data"
author:
- Damiano Oldoni
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

In this document we aggregate data. The goal is to produce a *data cube* at species level and its *taxonomic compendium*.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r load_libraries}
library(tidyverse)      # To do datascience
library(here)           # To find files
library(rgbif)          # To use GBIF services
library(glue)           # To write queries
library(RSQLite)        # To interact with SQlite databases
```

Define key returned in `1_download.Rmd` and country:

```{r define_key_countries}
key <- "0173465-210914110416597"
countries <- c("IT")
```

Name and path of `.sqlite` file:

```{r name_path}
sqlite_file <- paste0(paste(key, "occurrence", countries, sep = "_"), ".sqlite")
sqlite_path <- here::here("data", "interim", sqlite_file)
```

Table name:

```{r define_table_name}
table_name <- "occ"
```

Open connection to database:

```{r open_connection_to_db}
sqlite_occ <- dbConnect(SQLite(), dbname = sqlite_path)
```

# Aggregate at species level

## Aggregate occurrences

Species, synonyms of that species and infraspecific taxa of that species all share the same `speciesKey`.

For speeding up the aggregation, we create an index on `year`, `eea_cell_code` and `speciesKey` if not present:

```{r create_idx_speciesKey}
idx_species_year_cell <- "idx_species_year_cell"
# get indexes on table
query <- glue_sql(
    "PRAGMA index_list({table_name})",
    table_name = table_name,
    .con = sqlite_occ
)
indexes_all <- dbGetQuery(sqlite_occ, query)

# create index if not present
if (!idx_species_year_cell %in% indexes_all$name) {
  query <- glue_sql(
  "CREATE INDEX {`idx`} ON {table_name} ({`cols_idx`*})",
  idx = idx_species_year_cell,
  table_name = table_name,
  cols_idx = c("year", 
               "eea_cell_code", 
               "speciesKey",
               "coordinateUncertaintyInMeters"),
  .con = sqlite_occ
  )
  dbExecute(sqlite_occ, query)
}
```

Group by `year`, `eea_cell_code` and `speciesKey`, count number of occurrences for each group and get the lowest value of coordinate uncertainty `occurrenceUncertaintyInMeters`:

```{r get_datacube_species}
query <- glue_sql(
  "SELECT {`cols`*}, COUNT(_ROWID_), MIN({`coord_uncertainty`}) FROM {table} GROUP BY {`cols`*}",
  cols = c("year", 
           "eea_cell_code", 
           "speciesKey"),
  coord_uncertainty = "coordinateUncertaintyInMeters",
  table = table_name,
  .con = sqlite_occ
)
occ_cube_species <- 
  dbGetQuery(sqlite_occ, query) %>%
  rename(
    n = "COUNT(_ROWID_)",
    min_coord_uncertainty = "MIN(`coordinateUncertaintyInMeters`)") %>%
  mutate(speciesKey = as.numeric(speciesKey))
```

Preview:

```{r preview_occ_species}
occ_cube_species %>% head()
```

Number of occurrences linked to taxa with higher rank than species:

```{r occs_higher_rank_than_species}
occ_cube_species %>%
  filter(speciesKey == 0 | is.na(speciesKey)) %>%
  select(n) %>%
  colSums()
```

We will discard them:

```{r remove_occs_with_higher_rank}
occ_cube_species <-
  occ_cube_species %>%
  filter(speciesKey != 0 & !is.na(speciesKey))
```

## Map taxa

Grouping by `speciesKey`, we loose informations about which taxa share the same `speciesKey`. This information could be sometimes helpful. We extract it in a separate data.frame, `taxa_species`.

For speeding up the extraction, we create an index on `speciesKey`, `taxonKey` and `scientificName` if not present:

```{r create_idx_speciesKey}
idx_species_taxon_name <- "idx_species_taxon_name"
# get indexes on table
query <- glue_sql(
    "PRAGMA index_list({table_name})",
    table_name = table_name,
    .con = sqlite_occ
)
indexes_all <- dbGetQuery(sqlite_occ, query)

# create index if not present
if (!idx_species_taxon_name %in% indexes_all$name) {
  query <- glue_sql(
  "CREATE INDEX {`idx`} ON {table_name} ({`cols_idx`*})",
  idx = idx_species_taxon_name,
  table_name = table_name,
  cols_idx = c("speciesKey", 
               "taxonKey", 
               "scientificName"),
  .con = sqlite_occ
  )
  dbExecute(sqlite_occ, query)
}
```

We get first all distinct taxa:

```{r get_distinct_taxa_in_occ_cube_species}
query <- glue_sql(
  "SELECT DISTINCT {`cols`*} FROM {table}",
  cols = c("speciesKey",
           "taxonKey",
           "scientificName"),
  table = table_name,
  .con = sqlite_occ
)
occ_cube_species_taxa <- 
  dbGetQuery(sqlite_occ, query)
```

Taxa with rank higher than species whose occurrences we discarded:

```{r taxa_higher_rank_than_species}
occ_cube_species_taxa %>%
  filter(speciesKey == 0)
```

We remove them as well:

```{r remove_taxa_rank_higher_species}
occ_cube_species_taxa <- 
  occ_cube_species_taxa %>%
  filter(speciesKey != 0)
```

Some species have occurrences coming from multiple taxa, synonyms and/or infraspecific taxa:

```{r show_multiple_taxonKey}
occ_cube_species_taxa %>%
  group_by(speciesKey) %>%
  count() %>%
  filter(n > 1) %>%
  select(-n) %>%
  left_join(occ_cube_species_taxa, by = "speciesKey") %>%
  arrange(speciesKey, taxonKey)
```

Some species have occurrences only from taxa linked to their infraspecific taxa or synonyms. In these cases `speciesKey` is never equal to `taxonKey`:

```{r show_taxa_speciesKey_not_taxonKey}
occ_cube_species_taxa %>%
  group_by(speciesKey) %>%
  count() %>%
  rename(n_taxa = n) %>%
  left_join(occ_cube_species_taxa, by = "speciesKey") %>%
  group_by(speciesKey, n_taxa) %>%
  filter(taxonKey != speciesKey) %>%
  count() %>%
  rename(n_taxonKey_not_speciesKey = n) %>%
  filter(n_taxonKey_not_speciesKey == n_taxa) %>%
  left_join(occ_cube_species_taxa %>%
              filter(speciesKey != taxonKey),
            by = "speciesKey") %>%
  ungroup() %>%
  select(-c(n_taxa, n_taxonKey_not_speciesKey)) %>%
  arrange(speciesKey, taxonKey)
```

We create `taxa_species` by adding the taxonomic rank, `SPECIES`, and the taxonomic status of the species, one of `ACCEPTED` or `DOUBTFUL`, and create a column called `include` which contains all taxa whose occurrences are linked to the species:

```{r make_taxa_species}
taxa_species <- 
  occ_cube_species_taxa %>%
  
  # get unique 'speciesKey'
  distinct(speciesKey) %>%

  # extract speciesKey
  pull(speciesKey) %>%

  # GBIF query via name_usage
  map(~name_usage(key = .x)) %>%

  # Select data
  map(~.x[["data"]]) %>%

  # Merge all taxa in a data.frame
  reduce(full_join) %>%

  # select columns of interest
  select(speciesKey, scientificName, rank, taxonomicStatus, kingdom) %>%
  
  # rename 'scientificName' to 'species_scientificName'
  rename(species_scientificName = scientificName) %>%
  
  # add these columns to original df
  right_join(occ_cube_species_taxa, by = "speciesKey") %>%
  
  # group by 'speciesKey'
  group_by(speciesKey,
           species_scientificName,
           rank,
           taxonomicStatus,
           kingdom) %>%
  
  # create 'includes' column
  summarize(includes = paste(
    taxonKey, 
    scientificName, 
    sep = ": ", 
    collapse = " | ")) %>%
  
  # rename 'species_scientificName' to 'scientificName'
  rename(scientificName = species_scientificName)
taxa_species
```

# Save aggregated data

## Save aggregated occurrence data

Save the *occurrence data cube* as comma separated text file:

```{r save_cube}
occ_cube_species_filename <- paste0(
  paste(str_to_lower(countries), "species", "cube", sep = "_"),
  ".csv"
)
write_csv(occ_cube_species, 
          here::here("data",
                     "processed",
                     occ_cube_species_filename),
          na = ""
)
```

## Save taxa

Save the taxa as comma separated text file:

```{r save_cube_taxa}
taxa_species_filename <- paste0(
  paste(str_to_lower(countries), "species", "info", sep = "_"),
  ".csv"
)
write_csv(taxa_species,
          path = here::here("data",
                            "processed",
                            taxa_species_filename),
          na = "", 
          quote_escape = FALSE
)
```

Close connection:

```{r close_connection}
dbDisconnect(sqlite_occ)
```
